---
title: "TTC Bus Delay Analysis using Linear Predictive Model of Delay-Causing Incidents"
output:
  word_document: default
  html_notebook: default
  html_document:
    df_print: paged
  pdf_document: default
---


Load readxl & stringr libraries

```{r}
library(readxl)
library(stringr)
library(ggplot2)
require(coefplot)
library(tree)
library(rpart)
library(caret)
```

# Create one dataframe for the entire worksheet

```{r}
df <- data.frame()

for (b in c(14,15,16,17,18)){
    

    x <- sprintf("D:/DataAnalytics/CKME_136/TTC_Bus_Delay/Capstone/Bus_20%s.xlsx",b)

    
    y <- excel_sheets(x)
    
    
    # Create one dataframe for the entire worksheet
    
    for (a in y){
        
        df1 <- read_excel(x,sheet = a)
        
        df <- rbind(df,df1)
        
    }
    
}

```




# Store Combined file in one CSV file

```{r}

write.csv(df,"D:/DataAnalytics/CKME_136/TTC_Bus_Delay/Capstone/combined.csv")
```



# Load the combined file

```{r}
delay <- read.csv("D:/DataAnalytics/CKME_136/TTC_Bus_Delay/Capstone/combined.csv", header = TRUE, stringsAsFactors = FALSE, sep = ",")

head(delay)

delay$Incident <- as.factor(delay$Incident)

delay$Day <- as.factor(delay$Day)

class(delay$Incident)

class(delay$Day)

summary(delay$Incident)

```



# Data Cleaning

# Clean direction feature

```{r}
direct_fun <- function(x){
    
    switch(x,
           "W/N"="WNB","w/b"="WB","b/w"="BW","w.b"="WB","w"="WB","UP"="NB","u"="NB",
           "SS"="SB","s/d"="SB","s/b"="SB","s"="SB","Ou"="NB","W"="WB","N"="NB",
           "S"="SB","E"="EB","n"="NB","s"="SB","e"="EB","WB"="WB","NB"="NB","SB"="SB",
           "EB"="EB","BW"="BW","E/W"="BW","N/E"="NE","BW's"="BW","BWS"="BW","B/W's"="BW",
           "b/ws"="BW","b/b"="BW","'N"="NB", "N & S"="BW","1w/b"="WB"

    )
}

delay$Direction <- sapply(delay$Direction,direct_fun)


```


# Renaming the columns

```{r}
names(delay) <- c("SrNo","date","route","time","day","location","incident","delay","gap","direction","vehicle")

delay$direction <- as.character(delay$direction)
delay$direction <- as.factor(delay$direction)
```

# Removing "NULLs"

```{r}
delay <- delay[delay$direction != "NULL",]
```

# removing "NAs"

```{r}
delay <- delay[which(!is.na(delay$delay)),]


ggplot(delay,aes(y=delay,x=incident))+geom_boxplot()

```



# removing outliers

```{r}
delay <- delay[which(delay$delay < 60 & delay$delay > 0 ),]

head(delay)

class(delay$date)

ggplot(delay,aes(y=delay,x=incident))+geom_boxplot()


aggregate(delay ~ incident, data = delay, mean)

aggregate(delay ~ incident, data = delay, var)

ggplot(delay, aes(x= delay,fill=incident))+geom_histogram(binwidth = 3,alpha=3/4)+facet_wrap(~incident)

ggplot(delay, aes(x= delay,fill=incident))+geom_histogram(binwidth = 3,alpha=3/4)+facet_wrap(~incident)+theme(legend.position = "top")

```



# Analysis of Variance "ANOVA" using linear regression coefficient

```{r}
delay.lm <- lm(delay ~ incident, data = delay)
summary(delay.lm)


delay.lm2 <- lm(delay ~ incident-1, data = delay)
summary(delay.lm2)

```


# Plot linear regression coefficient

```{r}
coefplot(delay.lm)


coefplot(delay.lm2)

```



# Consider the interaction with the day of incident

```{r}
delay.lm3 <- lm(delay ~ incident * day, data = delay)
summary(delay.lm3)

coefplot(delay.lm3)

```



# Studying monthly distribution of the delay

```{r}
month.col <- str_sub(delay$date,start = 6, end = 7)

month.name <- function(x){
    
    switch(x,
           "01"="01Jan",
           "02"="02Feb",
           "03"="03Mar",
           "04"="04Apr",
           "05"="05May",
           "06"="06Jun",
           "07"="07Jul",
           "08"="08Aug",
           "09"="09Sep",
           "10"="10Oct",
           "11"="11Nov",
           "12"="12Dec"
           )
    
}

month.names <- sapply(month.col, month.name)

delay <- data.frame(delay,month.names)

ggplot(delay, aes(x= delay,fill=month.names))+geom_histogram(binwidth =3,alpha=3/4)+facet_wrap(~month.names)


yearly <- str_sub(delay$date,start = 1, end = 4)

delay <- data.frame(delay,yearly)

ggplot(delay, aes(x= delay,fill=incident))+geom_histogram(binwidth = 3,alpha=3/4)+facet_wrap(~yearly)

season <- function(x){
    
    switch(x,
           "01"="Winter",
           "02"="Winter",
           "03"="Winter",
           "04"="Spring",
           "05"="Spring",
           "06"="Spring",
           "07"="Summer",
           "08"="Summer",
           "09"="Fall",
           "10"="Fall",
           "11"="Fall",
           "12"="Winter"
    )
    
}
season.names <- sapply(month.col, season)

delay <- data.frame(delay,season.names)

ggplot(delay, aes(x= delay,fill=incident))+geom_histogram(binwidth = 3,alpha=3/4)+facet_wrap(~season.names)+theme(legend.position = "top")

ggplot(delay, aes(x= delay,fill=incident))+geom_histogram(binwidth = 3,alpha=3/4)+facet_wrap(~yearly+season.names)

```






# Create a prediction model for the delay through linear regression

```{r}

index <- sample(length(delay$route), .7 *length(delay$route))

train <- delay[index,]
test <- delay[-index,]


delay_single <- lm(delay ~ incident-1, data = train)

summary(delay_single)



delay_pred <- predict(delay_single, newdata = test, interval = "prediction")

pred <- data.frame(delay_pred)


error_pred <- pred$fit - test$delay

error <- data.frame(error_pred)
head(error)

hist(error_pred)

ggplot(error,aes(x=error_pred))+geom_histogram(binwidth = 3,alpha=3/4)

summary(error_pred)

mse_single <- sqrt(sum((error_pred)^2)/nrow(test))

mse_single

relative_error <- 1- ((test$delay - abs(error_pred))/test$delay)
#relative_error
relative_error_less25 <- table(relative_error < .25)["TRUE"]/nrow(test)
relative_error_less25_pct <- paste("relative_error_less 25% is ",100*relative_error_less25)
relative_error_less25_pct

```


    

# Hypothesis: delays due to "Mechanical" incidents has no significant difference from delays caused by "All incidents"

```{r}
Mechanical_mean <- mean(delay$delay[delay$incident == "Mechanical"])

Mechanical_mean

Mechanical_sd <- sd(delay$delay[delay$incident == "Mechanical"])

Mechanical_var <- var(delay$delay[delay$incident == "Mechanical"])



Population_mean <- mean(delay$delay)

Population_mean

Population_sd <- sd(delay$delay)

Population_sd

Population_var <- var(delay$delay)

Population_var

n <- length(delay$delay[delay$incident == "Mechanical"])

n

z <- (Mechanical_mean - Population_mean)/(Population_sd/sqrt(n))

z

z1 <- (Mechanical_mean - Population_mean)/(Mechanical_sd/sqrt(n))

p_value <- pnorm(z)

p_value

```




# side by side histograms of whole delays and delays due to Mechanical reasons

```{r}
ggplot(delay, aes(x=delay))+geom_histogram(binwidth = 3,alpha=3/4)


delay_mechanical <- delay[delay$incident == "Mechanical",]

ggplot(delay_mechanical, aes(x=delay))+geom_histogram(binwidth = 3,alpha=3/4)

```




# study Multi regression cosidering "gap" attribute
# Pearson correlation test between Delay and Gap

```{r}
delay.gap <- delay[!is.na(delay$gap), ]

cor.test(delay.gap$delay,delay.gap$gap, method = "pearson")

index.gap <- sample(length(delay.gap$route), .7 *length(delay.gap$route))

train.gap <- delay.gap[index.gap,]
test.gap <- delay.gap[-index.gap,]

delay_Multi.gap <- lm(delay ~ incident-1+gap, data = train.gap)

summary(delay_Multi.gap)



delay_pred.gap <- predict(delay_Multi.gap, newdata = test.gap, interval = "prediction")

pred_Multi.gap <- data.frame(delay_pred.gap)

error_pred_multi.gap <- pred_Multi.gap$fit - test.gap$delay

error_multi.gap <- data.frame(error_pred_multi.gap)
head(error_multi.gap)

sum(is.na(train.gap$gap))

hist(error_pred_multi.gap)

ggplot(error_multi.gap,aes(x=error_pred_multi.gap))+geom_histogram(binwidth = 3,alpha=3/4)

summary(error_pred_multi.gap)

mse_multi.gap <- sqrt(sum((error_pred_multi.gap)^2)/nrow(test.gap))

mse_multi.gap

```


# study Multi regression cosidering "direction" attribute

```{r}

delay_Multi.direction <- lm(delay ~ incident-1+direction, data = train)

summary(delay_Multi.direction)




delay_pred.direction <- predict(delay_Multi.direction, newdata = test, interval = "prediction")

pred_Multi.direction <- data.frame(delay_pred.direction)

error_pred_multi.direction <- pred_Multi.direction$fit - test$delay

error_multi.direction <- data.frame(error_pred_multi.direction)
head(error_multi.direction)

hist(error_pred_multi.direction)

ggplot(error_multi.direction,aes(x=error_pred_multi.direction))+geom_histogram(binwidth = 3,alpha=3/4)

mse_multi.direction <- sqrt(sum((error_multi.direction)^2)/nrow(test))

mse_multi.direction

```




# Consider "gap" & "direction" features

```{r}

delay_Multi.gap_direction <- lm(delay ~ incident-1+gap+direction, data = train.gap)

summary(delay_Multi.gap_direction)



delay_pred.gap_direction <- predict(delay_Multi.gap_direction, newdata = test.gap, interval = "prediction")

pred_Multi.gap_direction <- data.frame(delay_pred.gap_direction)

error_pred_multi.gap_direction <- pred_Multi.gap_direction$fit - test.gap$delay

error_multi.gap_direction <- data.frame(error_pred_multi.gap_direction)
head(error_multi.gap_direction)


hist(error_pred_multi.gap_direction)

ggplot(error_multi.gap_direction,aes(x=error_pred_multi.gap_direction))+geom_histogram(binwidth = 3,alpha=3/4)

summary(error_pred_multi.gap_direction)

mse_multi.gap_direction <- sqrt(sum((error_pred_multi.gap_direction)^2)/nrow(test.gap))

mse_multi.gap_direction

```


# Create a prediction model for the delay through regression tree
# consider "incident" feature 

```{r}
tree_model <- tree(delay ~ incident, data = test)
plot(tree_model)
text(tree_model)

class(delay$direction)
levels(delay$direction)

delay_pred_tree <- predict(tree_model, newdata = test, interval = "prediction")

pred_tree <- data.frame(delay_pred_tree)

error_pred_tree <- pred_tree$delay_pred_tree - test$delay

error_tree <- data.frame(error_pred_tree)
head(error_tree)

hist(error_pred_tree)


mse_tree <- sqrt(sum((error_tree)^2)/nrow(test))

mse_tree

```

# Create a prediction model for the delay through regression tree
# consider "incident" & "direction" features


```{r}
tree_model_multi <- tree(delay ~ incident+direction, data = test)
plot(tree_model_multi)
text(tree_model_multi)

delay_pred_tree_multi <- predict(tree_model_multi, newdata = test, interval = "prediction")

pred_tree_multi <- data.frame(delay_pred_tree_multi)

error_pred_tree_multi <- pred_tree_multi$delay_pred_tree_multi - test$delay

error_tree_multi <- data.frame(error_pred_tree_multi)
head(error_tree_multi)

hist(error_pred_tree_multi)


mse_tree_multi <- sqrt(sum((error_tree_multi)^2)/nrow(test))

mse_tree_multi


```


# Create a prediction model for the delay through regression tree
# consider "incident", "gap" features


```{r}
tree_model_multi.gap <- tree(delay ~ incident+gap, data = test.gap)
plot(tree_model_multi.gap)
text(tree_model_multi.gap)

delay_pred_tree_multi.gap <- predict(tree_model_multi.gap, newdata = test.gap, interval = "prediction")

pred_tree_multi.gap <- data.frame(delay_pred_tree_multi.gap)

error_pred_tree_multi.gap <- pred_tree_multi.gap$delay_pred_tree_multi.gap - test.gap$delay

error_tree_multi.gap <- data.frame(error_pred_tree_multi.gap)
head(error_tree_multi.gap)

hist(error_pred_tree_multi.gap)


mse_tree_multi.gap <- sqrt(sum((error_tree_multi.gap)^2)/nrow(test.gap))

mse_tree_multi.gap


```


# Cross Validation 10 folds

# using incidents and directions

```{r}
# Define training control
set.seed(123) 
train.control <- trainControl(method = "cv", number = 10)
# Train the model
model <- train(delay ~ incident+direction, data = delay, method = "lm", trControl = train.control)
# Summarize the results
print(model)

```

# Cross Validation 10 folds

# using incidents

```{r}
# Define training control
set.seed(123) 
train.control <- trainControl(method = "repeatedcv", number = 10,repeats = 3)
# Train the model
model <- train(delay ~ incident, data = delay, method = "lm", trControl = train.control)
# Summarize the results
print(model)

```


# Cross Validation 10 folds

# using incidents and gap

```{r}
# Define training control
set.seed(123) 
train.control <- trainControl(method = "repeatedcv", number = 10,repeats = 3)
# Train the model
model <- train(delay ~ incident+gap, data = delay.gap, method = "lm", trControl = train.control)
# Summarize the results
print(model)

```


# Cross Validation 3 folds

# using incidents

```{r}
# Define training control
set.seed(123) 
train.control <- trainControl(method = "cv", number = 3)
# Train the model
model <- train(delay ~ incident, data = delay, method = "M5", trControl = train.control)
# Summarize the results
print(model)

```



# Cross Validation 3 folds

# using incidents

```{r}
# Define training control
set.seed(123) 
train.control <- trainControl(method = "cv", number = 3)
# Train the model
model.gap <- train(delay ~ incident+gap, data = delay.gap, method = "M5", trControl = train.control)
# Summarize the results
print(model.gap)

```